Transfer learning between Sentinel-1 acquisition modes enhances \\ the few-shot segmentation of natural oil slicks in the Arctic

### 2025-03-31 latex/manuscript.tex
ref update

Researchers recently documented two seeps in the Norwegian waters. First, Panieri et al. \cite{panieriArcticNaturalOil2024} investigated an intermittent seep offshore Prins Karls Forland 
(10.42°E, 78.49°N), west of the Svalbard archipelago. At this location, we manually delineated oil slicks extending from a common seeping point, in 70 images between 2015 and 2024.
Second, Serov et al. \cite{serovWidespreadNaturalMethane2023} (but see also Ivanov et al. \cite{ivanovSearchDetectionNatural2020}) gave extensive evidence of a wide seepage area in the Barents Sea. 
This area is part of the Sentralbanken high, a geological structure with prospective hydrocarbon potential, although it remains largely unexplored \cite{lundschienNorthBarentsComposite2025}. 
There (31.33-32.37°E, 75.15°N-75.35°N), we mapped 250 natural slicks in 88 images between 2015 and 2020.
At those two locations, labeled samples from the horizontal co-polarization (HH) band of Sentinel-1 EW images make the $\mathit{D}_T$ (Fig. \ref{figure1}a, \ref{figure1}b). 
At lower latitudes, oil slicks of various kinds can be observed in the North Sea, an hydrocarbon-rich and seepage-active region \cite{juddSeabedFluidFlow2009,hovlandSeabedPockmarksSeepages1988}. 
With the same procedure, we mapped over 1700 slicks, this time using the vertical co-polarization (VV) band of 463 Sentinel-1 IW images. Here, labeled samples in the North Sea make the $\mathit{D}_S$ (Fig. \ref{figure1}c).
Only the co-polarization bands were used, considering their oil-discrimination capacity \cite{johanssonMultifrequencyPolarimetricSAR2017,brekkeSAROilSpill2020}.
Since a slick can be split in more than one part if its extent is not fully connected, we give slick numbers in Table \ref{table1} as approximations only.
\begin{figure}[!t]
    \centering
    \includegraphics[width=3.3in]{figures/layout2_grids.png} %3.3=left 3.5=right
    \caption{a) Target domain \( \mathit{D}_T \): Mapped slicks offshore Prins Karls Forland, Svalbard \quad b) Target domain \( \mathit{D}_T \): Mapped slicks in Sentralbanken high, Barents Sea \quad 
    c) Source domain \( \mathit{D}_S \): Mapped slicks in the North Sea}
    \label{figure1}
\end{figure}
\begin{table}[!t]
    \caption{Labeled Slicks in the Source and Target Domains}
    \label{table1}
    \centering
    \begin{tabular}{lcc|cc}
        \toprule
        & \multicolumn{2}{c}{\textbf{Source domain}} & \multicolumn{2}{c}{\textbf{Target domain}} \\
        \cmidrule(lr){2-3} \cmidrule(lr){4-5}
        \textbf{Sentinel-1 mode/pol.} & \multicolumn{2}{c}{IW / VV} & \multicolumn{2}{c}{EW / HH} \\
        \midrule
        \textbf{Area} & \multicolumn{2}{c}{North Sea} & Svalbard & Barents Sea \\
        \textbf{Number of images} & \multicolumn{2}{c}{463} & 70 & 88 \\
        \textbf{Period} & \multicolumn{2}{c}{2017-2021} & 2015-2024 & 2015-2020 \\
        \textbf{Number of slicks} & \multicolumn{2}{c}{$\sim 1723$} & $\sim 81$ & $\sim 250$ \\
        \bottomrule
    \end{tabular}
\end{table}

>>>>

Researchers recently documented two seeps in the Norwegian waters. First, Panieri et al. \cite{panieriArcticNaturalOil2024} investigated an intermittent seep offshore Prins Karls Forland 
(10.42°E, 78.49°N), west of the Svalbard archipelago. At this location, we manually delineated oil slicks extending from a common seeping point, in 70 images between 2015 and 2024.
Second, Serov et al. \cite{serovWidespreadNaturalMethane2023} (but see also Ivanov et al. \cite{ivanovSearchDetectionNatural2020}) gave extensive evidence of a wide seepage area in the Barents Sea. 
This area is part of the Sentralbanken high, a geological structure with prospective hydrocarbon potential, although it remains largely unexplored \cite{lundschienNorthBarentsComposite2025}. 
There (31.33-32.37°E, 75.15°N-75.35°N), we mapped 250 natural slicks in 88 images between 2015 and 2020.
At those two locations, labeled samples from the horizontal co-polarization (HH) band of Sentinel-1 EW images make the $\mathit{D}_T$ (Fig. \ref{figure1}a, \ref{figure1}b). 
At lower latitudes, oil slicks of various kinds can be observed in the North Sea, an hydrocarbon-rich and seepage-active region \cite{hovlandChapter2Focus1988}. 
With the same procedure, we mapped over 1700 slicks, this time using the vertical co-polarization (VV) band of 463 Sentinel-1 IW images. Here, labeled samples in the North Sea make the $\mathit{D}_S$ (Fig. \ref{figure1}c).
Only the co-polarization bands were used, considering their oil-discrimination capacity \cite{kudryavtsevDualCoPolarizedSAR2013,johanssonMultifrequencyPolarimetricSAR2017,brekkeSAROilSpill2020}.
Since a slick can be split in more than one part if its extent is not fully connected, we give slick numbers in Table \ref{table1} as approximations only.
\begin{figure}[!t]
    \centering
    \includegraphics[width=3.3in]{figures/layout2_grids.png} %3.3=left 3.5=right
    \caption{a) Target domain \( \mathit{D}_T \): Mapped slicks offshore Prins Karls Forland, Svalbard \quad b) Target domain \( \mathit{D}_T \): Mapped slicks in Sentralbanken high, Barents Sea \quad 
    c) Source domain \( \mathit{D}_S \): Mapped slicks in the North Sea}
    \label{figure1}
\end{figure}
\begin{table}[!t]
    \caption{Labeled Slicks in the Source and Target Domains}
    \label{table1}
    \centering
    \begin{tabular}{lcc|cc}
        \toprule
        & \multicolumn{2}{c}{\textbf{Source domain}} & \multicolumn{2}{c}{\textbf{Target domain}} \\
        \cmidrule(lr){2-3} \cmidrule(lr){4-5}
        \textbf{Sentinel-1 mode/pol.} & \multicolumn{2}{c}{IW / VV} & \multicolumn{2}{c}{EW / HH} \\
        \midrule
        \textbf{Area} & \multicolumn{2}{c}{North Sea} & Svalbard & Barents Sea \\
        \textbf{Number of images} & \multicolumn{2}{c}{463} & 70 & 88 \\
        \textbf{Period} & \multicolumn{2}{c}{2017-2021} & 2015-2024 & 2015-2020 \\
        \textbf{Number of slicks} & \multicolumn{2}{c}{$\sim 1723$} & $\sim 81$ & $\sim 250$ \\
        \bottomrule
    \end{tabular}
\end{table}
